---
title: "Individual, Population, and Conservation Summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Individual, Population, and Conservation Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, message=FALSE, warning=FALSE}
library(Shearwater)
library(dplyr)
library(geosphere)
library(sf)
```

## Overview

This vignette picks up at the end of the movement-metrics pipeline. By this
point your data have been read in, cleaned, and segmented into foraging trips
(Vignettes 1–3). Here we:

1. Aggregate per-trip metrics to **individual** and **population** level.
2. Estimate each bird's **foraging range** as a spatial polygon.
3. **Map** the raw tracks.
4. **Export** summary tables and GIS layers for reporting.

Standard column names used throughout the package:

| Concept | Column name |
|---|---|
| Individual ID | `track_id` |
| Timestamp (UTC) | `datetime_gmt` |
| Longitude | `longitude` |
| Latitude | `latitude` |
| At-colony flag | `at_colony` |
| Trip ID | `trip_id` |
| Behavioural phase | `phase` |

---

## 1  Load and standardise the example data

```{r load-data}
gps_path <- system.file(
  "extdata",
  "example_wtsh_data.csv",
  package = "Shearwater",
  mustWork = TRUE
)

raw <- read.csv(
  gps_path,
  stringsAsFactors = FALSE
)

tracks <- raw %>%
  mutate(
    track_id = "bird_001",
    datetime_gmt = as.POSIXct(
      paste(Date, Time),
      format = "%m/%d/%Y %H:%M:%S",
      tz = "UTC"
    ),
    longitude = Longitude,
    latitude = Latitude
  ) %>%
  arrange(track_id, datetime_gmt)

# Colony location — use first GPS point for this example
colony <- c(
  lon = tracks$longitude[1],
  lat = tracks$latitude[1]
)

tracks <- tracks %>%
  mutate(
    dist_to_colony_m = geosphere::distHaversine(
      cbind(longitude, latitude),
      matrix(
        c(colony["lon"], colony["lat"]),
        nrow = n(),
        ncol = 2,
        byrow = TRUE
      )
    ),
    at_colony = dist_to_colony_m <= 75
  )

head(tracks)
```

---

## 2  Build the per-trip metrics table

Cleaning, segmentation, and individual metric calculation are covered in
Vignettes 2–3. `summarize_movement()` runs the full pipeline in one call and
returns a named list containing every intermediate result.

```{r run-pipeline, eval=FALSE}
results <- summarize_movement(
  track           = tracks,
  colony_coords   = colony,
  bird_id_col     = "track_id",
  datetime_col    = "datetime_gmt",
  lon_col         = "longitude",
  lat_col         = "latitude",
  colony_flag_col = "at_colony",
  duration_units  = "hours",
  include_spatial = TRUE
)

# Unpack the results list
trip_metrics       <- results$trip_metrics
individual_metrics <- results$individual_metrics
population_metrics <- results$population_metrics
foraging_ranges    <- results$foraging_ranges
```

For the rest of this vignette we use a representative mock table built from
realistic Wedge-tailed Shearwater values so the code runs without the full
upstream pipeline.

```{r mock-trip-metrics}
# Distances / path lengths in metres; duration in hours (set by duration_units).
trip_metrics <- tibble::tribble(
  ~track_id,  ~trip_id, ~trip_duration, ~trip_distance_m,
              ~path_length_m, ~max_distance_from_colony_m,
  "WTSH_001", 1L,  33.6,  312400,  485200,  68300,
  "WTSH_001", 2L,  38.8,  445200,  621000,  89400,
  "WTSH_002", 1L,  18.0,  189300,  285100,  42100,
  "WTSH_002", 2L,  36.0,  398700,  591200,  78900,
  "WTSH_002", 3L,  21.0,  215600,  318400,  51300,
  "WTSH_003", 1L,  18.0,  178400,  261500,  38700,
  "WTSH_004", 1L,  42.0,  421500,  628900,  85200,
  "WTSH_004", 2L,  38.0,  387200,  578100,  79600,
  "WTSH_005", 1L,  18.0,  195800,  291400,  45200
)

glimpse(trip_metrics)
```

---

## 3  Individual-level summaries — `summarize_individual_metrics()`

`summarize_individual_metrics()` collapses the per-trip table to **one row per
bird**. For each metric column it computes the mean, maximum, and running total
across that bird's trips.

```{r individual-summary, eval=FALSE}
individual_metrics <- summarize_individual_metrics(
  trip_metrics     = trip_metrics,
  bird_id_col      = "track_id",
  trip_id_col      = "trip_id",
  distance_col     = "trip_distance_m",
  duration_col     = "trip_duration",
  path_length_col  = "path_length_m",
  max_distance_col = "max_distance_from_colony_m"
)
```

```{r individual-summary-mock, echo=FALSE}
# Mirrors the output of summarize_individual_metrics() exactly.
individual_metrics <- trip_metrics |>
  group_by(track_id) |>
  summarise(
    n_trips                         = n_distinct(trip_id),
    mean_trip_distance_m            = mean(trip_distance_m,           na.rm = TRUE),
    max_trip_distance_m             = max(trip_distance_m,            na.rm = TRUE),
    total_trip_distance_m           = sum(trip_distance_m,            na.rm = TRUE),
    mean_trip_duration              = mean(trip_duration,             na.rm = TRUE),
    max_trip_duration               = max(trip_duration,              na.rm = TRUE),
    total_trip_duration             = sum(trip_duration,              na.rm = TRUE),
    mean_path_length_m              = mean(path_length_m,             na.rm = TRUE),
    max_path_length_m               = max(path_length_m,              na.rm = TRUE),
    total_path_length_m             = sum(path_length_m,              na.rm = TRUE),
    mean_max_distance_from_colony_m = mean(max_distance_from_colony_m, na.rm = TRUE),
    max_distance_from_colony_m      = max(max_distance_from_colony_m,  na.rm = TRUE),
    .groups = "drop"
  )
```

A readable summary of the key metrics (converted to km for display):

```{r show-individual}
individual_metrics |>
  transmute(
    track_id,
    n_trips,
    mean_trip_dist_km    = round(mean_trip_distance_m / 1000, 1),
    max_trip_dist_km     = round(max_trip_distance_m  / 1000, 1),
    mean_trip_dur_hr     = round(mean_trip_duration, 1),
    max_trip_dur_hr      = round(max_trip_duration, 1),
    mean_max_colony_km   = round(mean_max_distance_from_colony_m / 1000, 1)
  )
```

### What the columns mean

| Column | Description |
|---|---|
| `n_trips` | Number of complete foraging trips recorded |
| `mean_trip_distance_m` | Mean straight-line distance per trip |
| `max_trip_distance_m` | Longest single trip |
| `total_trip_distance_m` | Cumulative distance across all trips |
| `mean_trip_duration` | Mean trip length (units set by `duration_units`) |
| `max_trip_duration` | Longest single trip |
| `total_trip_duration` | Total time away from colony |
| `mean_path_length_m` | Mean cumulative track length per trip |
| `mean_max_distance_from_colony_m` | How far out the bird typically goes |
| `max_distance_from_colony_m` | Furthest point ever reached |

### Interpreting individual variation

```{r individual-range}
individual_metrics |>
  summarise(
    dist_range_km  = paste0(
      round(min(mean_trip_distance_m) / 1000, 0), " – ",
      round(max(mean_trip_distance_m) / 1000, 0), " km"),
    dur_range_hr   = paste0(
      round(min(mean_trip_duration), 1), " – ",
      round(max(mean_trip_duration), 1), " hr"),
    radius_range_km = paste0(
      round(min(mean_max_distance_from_colony_m) / 1000, 0), " – ",
      round(max(mean_max_distance_from_colony_m) / 1000, 0), " km")
  )
```

Even in this five-bird sample, mean trip distance spans a nearly two-fold range.
Birds making longer, farther trips may be accessing distant prey patches or may
have fewer constraints on time away from the nest (e.g. failed breeders). Always
check breeding-status metadata before drawing ecological conclusions from
individual variation.

---

## 4  Population-level summaries — `summarize_population_metrics()`

`summarize_population_metrics()` takes `individual_metrics` and produces a
**single-row colony snapshot**. It auto-detects every numeric column and
computes mean, SD, min, and max for each, appending the statistic as a suffix.

```{r population-summary, eval=FALSE}
population_metrics <- summarize_population_metrics(
  individual_metrics = individual_metrics,
  bird_id_col        = "track_id"
)
```

```{r population-summary-mock, echo=FALSE}
numeric_cols <- names(individual_metrics)[sapply(individual_metrics, is.numeric)]

population_metrics <- bind_cols(
  summarise(individual_metrics, n_individuals = n_distinct(track_id)),
  summarise(individual_metrics,
    across(
      all_of(numeric_cols),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd   = ~ sd(.x,   na.rm = TRUE),
        min  = ~ min(.x,  na.rm = TRUE),
        max  = ~ max(.x,  na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )
)
```

The raw output is wide (mean/SD/min/max for every individual-level column).
Select the columns most relevant to your reporting context:

```{r show-population}
population_metrics |>
  select(
    n_individuals,
    starts_with("n_trips"),
    starts_with("mean_trip_distance_m"),
    starts_with("mean_trip_duration"),
    starts_with("mean_max_distance_from_colony_m")
  ) |>
  mutate(across(contains("_m_"), ~ round(.x / 1000, 1))) |>  # convert m → km
  glimpse()
```

### Reporting-ready text

```{r population-prose}
cat(sprintf(
  paste0(
    "Across %d individuals (%d total trips), Wedge-tailed Shearwaters at\n",
    "Kaena Point made foraging trips lasting %.1f \u00b1 %.1f hr (mean \u00b1 SD)\n",
    "and covering %.0f \u00b1 %.0f km. On average, birds reached a maximum\n",
    "of %.0f \u00b1 %.0f km from the colony."
  ),
  population_metrics$n_individuals,
  population_metrics$n_trips_mean * population_metrics$n_individuals,
  population_metrics$mean_trip_duration_mean,
  population_metrics$mean_trip_duration_sd,
  population_metrics$mean_trip_distance_m_mean / 1000,
  population_metrics$mean_trip_distance_m_sd   / 1000,
  population_metrics$mean_max_distance_from_colony_m_mean / 1000,
  population_metrics$mean_max_distance_from_colony_m_sd   / 1000
))
```

> **Sample size note.** Population-level means are only as reliable as the
> individuals behind them. Aim for ≥ 10 tracked birds before presenting
> colony-level conclusions, and inspect `individual_metrics` for outliers first.

---

## 5  Estimating foraging range — `calc_foraging_range()`

`calc_foraging_range()` computes the spatial extent of each bird's foraging
activity, returning an `sf` polygon object. Two methods are available:

- `"convex_hull"` — minimum polygon enclosing all foraging points (default).
- `"bbox"` — axis-aligned bounding box; faster but always rectangular.

If `phase_col` is supplied, only points labelled with `foraging_value` are
included. Without a phase column, all track points are used.

```{r calc-foraging-range, eval=FALSE}
# Requires classify_trip_phase() to have been run first (adds the phase column)
foraging_ranges <- calc_foraging_range(
  track_data     = processed_track,
  bird_id_col    = "track_id",
  lon_col        = "longitude",
  lat_col        = "latitude",
  phase_col      = "phase",          # omit to use all fixes
  foraging_value = "foraging",
  method         = "convex_hull",
  crs            = 4326,
  area_crs       = 3857              # projected CRS used for area calculation
)
```

```{r mock-foraging-ranges, echo=FALSE}
# Representative convex-hull areas for WTSH foraging around Oahu (km²)
foraging_ranges <- tibble::tibble(
  track_id           = c("WTSH_001","WTSH_002","WTSH_003","WTSH_004","WTSH_005"),
  n_points           = c(1472L, 1504L, 361L, 1603L, 372L),
  foraging_range_km2 = c(4821.3, 2914.7, 1823.5, 4412.8, 2156.2)
)
```

```{r show-foraging-ranges}
foraging_ranges |>
  arrange(desc(foraging_range_km2)) |>
  mutate(foraging_range_km2 = round(foraging_range_km2, 0))
```

### Joining foraging range to the individual summary

Because `calc_foraging_range()` returns one row per `track_id`, it joins
cleanly onto `individual_metrics`:

```{r join-ranges}
individual_metrics |>
  left_join(
    foraging_ranges |> select(track_id, foraging_range_km2),
    by = "track_id"
  ) |>
  transmute(
    track_id,
    n_trips,
    mean_max_colony_km = round(mean_max_distance_from_colony_m / 1000, 1),
    foraging_range_km2 = round(foraging_range_km2, 0)
  ) |>
  arrange(desc(foraging_range_km2))
```

Birds with larger foraging ranges tend to have higher maximum colony distances,
but the convex hull also expands when a bird visits many spatially separated
areas within the same overall radius — making range area a complementary metric
to maximum distance rather than a substitute for it.

---

## 6  Visualising tracks — `plot_tracks()`

A track map is the quickest sanity check after running the pipeline and a
practical communication tool for non-technical audiences.

> **Column name note.** `plot_tracks()` expects columns named `lon` and `lat`.
> Rename them from the package standard (`longitude` / `latitude`) before
> plotting.

```{r plot-tracks, fig.width=6.5, fig.height=5.5,
    fig.alt="Map of WTSH GPS tracks coloured by individual, with colony marked."}
tracks |>
  rename(lon = longitude, lat = latitude) |>
  plot_tracks(
    color_by      = "track_id",
    colony_coords = colony,
    title         = "Wedge-tailed Shearwater foraging tracks\nKaena Point, Oahu — August 2013"
  )
```

For a closer look at a specific trip, use `plot_trip_map()` (see the
trip-segmentation vignette):

```{r plot-trip, eval=FALSE}
cleaned |>
  rename(lon = longitude, lat = latitude) |>
  plot_trip_map(
    trip_ids      = 1L,
    colony_coords = colony,
    title         = "WTSH_001 — Trip 1"
  )
```

---

## 7  Exporting results

### 7.1  Summary tables — `export_policy_summary_tables()`

`export_policy_summary_tables()` writes summary data frames to CSV for reports,
databases, or collaborators who do not use R.

```{r export-tables, eval=FALSE}
# Individual-level summary (one row per bird)
export_policy_summary_tables(
  data      = individual_metrics,
  file_path = "outputs/wtsh_individual_metrics.csv"
)

# Population-level snapshot (one row)
export_policy_summary_tables(
  data      = population_metrics,
  file_path = "outputs/wtsh_population_metrics.csv"
)
```

Both files are plain CSV and can be opened in Excel or imported into any
statistical software. The population summary is compact enough to include
directly in a species assessment; the individual summary supports the
underlying statistical analyses.

### 7.2  Spatial layers — `export_gis_layers()`

`export_gis_layers()` writes `sf` objects to GIS-ready formats (GeoPackage by
default). Use this to share foraging-range polygons with GIS analysts or to
load them into QGIS or ArcGIS for overlay with management boundaries.

```{r export-gis, eval=FALSE}
# Export foraging-range convex hulls
export_gis_layers(
  spatial_object = foraging_ranges,
  file_path      = "outputs/wtsh_foraging_ranges.gpkg"
)

# Export cleaned track points
export_gis_layers(
  spatial_object = sf::st_as_sf(
    cleaned,
    coords = c("longitude", "latitude"),
    crs    = 4326
  ),
  file_path = "outputs/wtsh_tracks.gpkg"
)
```

> **CRS check.** `export_gis_layers()` calls `assert_crs()` internally. If your
> `sf` object has no CRS assigned, set one before exporting:
> `sf::st_crs(foraging_ranges) <- 4326`

---

## Summary of functions used

| Function | Input | Output |
|---|---|---|
| `summarize_individual_metrics()` | One-row-per-trip metrics table | One-row-per-bird: mean, max, total of each metric |
| `summarize_population_metrics()` | Individual metrics table | Single row: mean/SD/min/max of all individual metrics |
| `calc_foraging_range()` | Track points (+ optional phase column) | `sf` polygon per bird with `foraging_range_km2` |
| `plot_tracks()` | Track data with `lon`/`lat` columns | `ggplot2` map |
| `export_policy_summary_tables()` | Any summary data frame | CSV file on disk |
| `export_gis_layers()` | `sf` object | GeoPackage / shapefile on disk |

## Next steps

- **Kernel utilization distributions**: use `estimate_kernel_ud()`,
  `get_core_area()`, and `get_home_range()` for a probabilistic alternative to
  the convex-hull foraging range (Vignette 5).
- **Fisheries overlap**: pass `foraging_ranges` and track data into
  `analyze_fisheries_overlap()` to quantify bycatch exposure by gear type
  (Vignette 6).
- **Jurisdiction overlay**: run `analyze_jurisdiction_overlap()` to determine
  how much foraging time falls inside EEZs versus the high seas (Vignette 7).
